### 2025-08-19 2025/Problems of Informatics/article.tex
last fixes

From a mathematical point of view, such a task can be represented as the problem of finding a global minimum of a function, given in the form of a ``black box'' and computable only in some part of the search domain. In this case, each computation of the objective function value at some point of the feasible domain may require significant computing resources. The objective function is assumed to satisfy the Lipschitz condition. The existence of subdomains where the objective function is undefined can be interpreted as the existence of some hidden, a priori unknown constraints of the problem. 

@@

%13
\item \label{rfa:enlit:Birect2020}
Paulavi{\v c}ius~R., Sergeyev~Y.D., Kvasov~D.E., {\v Z}ilinskas~J. Globally-biased {BIRECT} algorithm with local accelerators for expensive global optimization~// 
Expert Syst. Appl. 2020. Vol.~144. p.~113052. %DOI: 10.1016/j.eswa.2019.113052.

@@

%21
\item \label{rfa:enlit:Strongin2020}
Strongin~R.G., Barkalov~K.A., Bevzuk~S.A. Global optimization method with dual Lipschitz constant estimates for problems with non-convex constraints~// 
Soft Comput. 2020. Vol.~24, No.~16. P.~11853--11865. %DOI: 10.1007/s00500-020-05078-1.

@@

%%10
%\item \label{rfa:enlit:Kvasov2013}
%Kvasov~D.E., Sergeev~Ya.D. Metody lipshicevoj globalnoj optimizacii v zadachax upravleniya~// Avtomatika i telemexanika. 2013. No.~9. S.~3--19.
\item \label{rfa:enlit:Usova2024}
Usova~M.A., Barkalov~K.A. An Algorithm for Finding the Global Extremum of a Partially Defined Function~// Communications in Computer and Information Science. 2024. Vol.~1914. P.~147--161. %DOI: 10.1007/978-3-031-52470-7{\_}13.

@@

\item \label{rfa:enlit:iOptPaper}
Sysoev~A.V., Kozinov~E.A., Barkalov~K.A., Lebedev~I.G., Karchkov~D.A., Rodionov~D.M. Frejmvork metodov intellektualnoj evristicheskoj optimizacii iOpt // V kn.: Superkompyuternye dni v Rossii: Trudy mezhdunarodnoj konferencii. 2023. S.~179--185.

\item \label{rfa:enlit:iOptGithub}
Isxodnyj kod frejmvorka iOpt -- URL: https://github.com/aimclub/iOpt (data obrashheniya: 26.01.2025).

\item \label{rfa:enlit:iOptDocs}
Dokumentaciya iOpt -- URL: https://iopt.readthedocs.io/ru/latest/ (data obrashheniya: 26.01.2025).

@@

\item \label{rfa:enlit:nikitin2022automated}
Filippou~K., Aifantis~G., Papakostas~G.A., Tsekouras~G.E. Structure learning and hyperparameter optimization using an automated machine learning ({AutoML}) pipeline // Information. 2023. Vol.~14, No. 4. p.~232.

\item \label{rfa:enlit:MonthlyBeerArticle}
Xu~N. Time Series Analysis on Monthly Beer Production in Australia~// Highlights in Science, Engineering and Technology. 2024. Vol.~94. P.~392--401. DOI: 10.54097/4z3krj13

@@

%\item \label{rfa:enlit:FEDOT}
%Automated modeling and machine learning framework {FEDOT} -- URL:  https://github.com/aimclub/FEDOT (data obrashheniya: 25.07.2025).

@@

@@

В качестве второй модельной задачи рассмотрим задачу настройки гиперпараметров методов предсказания значений временного ряда (time series forecasting). Предсказание значений производилось с использованием фреймворка FEDOT [\ref{rfa:rulit:nikitin2022automated}]
(фреймворк с открытым исходным кодом для задач автоматизированного моделирования и машинного обучения, AutoML). FEDOT поддерживает полный цикл решения задачи, включающий препроцессинг, выбор модели, настройку, кроссвалидацию и т.д. Фреймворк использует ML-модели, в основном из стандартных библиотек \textit{sklearn}, \textit{statsmodels} и \textit{keras}.

@@

Для сравнения поставленная задача также была решена при помощи известного фреймворка оптимизации гиперпараметров Optuna%\cite{OptunaURL,optuna}
. Оба решателя были интегрированы в FEDOT и задействованы в вычислительном эксперименте.

@@

\item \label{rfa:rulit:Frazier2018}
Frazier~P.I. A Tutorial on Bayesian Optimization~// arXiv. 2018. DOI: 10.48550/arXiv.1807.02811.

@@

%13
\item \label{rfa:rulit:Birect2020}
Paulavi{\v c}ius~R., Sergeyev~Y.D., Kvasov~D.E., {\v Z}ilinskas~J. Globally-biased {BIRECT} algorithm with local accelerators for expensive global optimization~// Expert Syst. Appl. 2020. Vol.~144. p.~113052. DOI: 10.1016/j.eswa.2019.113052.

@@

%1
\item \label{rfa:rulit:Audet2022}
Audet~C., Batailly~A., Kojtych~S. Escaping unknown discontinuous regions in blackbox optimization~// SIAM J. Optim. 2022. Vol.~32, No.~3. P.~1843--1870. DOI: 10.1137/21m1420915.

%3
\item \label{rfa:rulit:Candelieri2019}
Candelieri~A. Sequential model based optimization of partially defined functions
under unknown constraints~// J. Glob. Optim. 2019. Vol.~79, No.~2. P.~281--303. DOI: 10.1007/s10898-019-00860-4.

@@

\item \label{rfa:rulit:Usova2024}
Usova~M.A., Barkalov~K.A. An Algorithm for Finding the Global Extremum of a Partially Defined Function~//
Communications in Computer and Information Science. 2024. Vol.~1914. P.~147--161. DOI: 10.1007/978-3-031-52470-7{\_}13.

@@

\item \label{rfa:rulit:nikitin2022automated}
Filippou~K., Aifantis~G., Papakostas~G.A., Tsekouras~G.E. Structure learning and hyperparameter optimization using an automated machine learning ({AutoML}) pipeline // Information. 2023. Vol.~14, No. 4. p.~232.

@@

%\item \label{rfa:rulit:fedot}
%Automated modeling and machine learning framework {FEDOT} -- URL:  https://github.com/aimclub/FEDOT (дата обращения: 25.07.2025).

>>>>

From a mathematical point of view, such a task can be re\-pre\-sen\-ted as the problem of finding a global minimum of a function, given in the form of a ``black box'' and computable only in some part of the search domain. In this case, each computation of the objective function value at some point of the feasible domain may require significant computing resources. The objective function is assumed to satisfy the Lipschitz condition. The existence of subdomains where the objective function is undefined can be interpreted as the existence of some hidden, a priori unknown constraints of the problem. 

@@

%13
\item \label{rfa:enlit:Birect2020}
Paulavi{\v c}ius~R., Sergeyev~Y.D., Kvasov~D.E., {\v Z}ilinskas~J. Globally-biased {BIRECT} al\-go\-rithm with local accelerators for expensive global optimization~// 
Expert Syst. Appl. 2020. Vol.~144. p.~113052. %DOI: 10.1016/j.eswa.2019.113052.

@@

%21
\item \label{rfa:enlit:Strongin2020}
Strongin~R.G., Barkalov~K.A., Bevzuk~S.A. Global optimization method with dual Lip\-schitz constant estimates for problems with non-convex constraints~// 
Soft Comput. 2020. Vol.~24, No.~16. P.~11853--11865. %DOI: 10.1007/s00500-020-05078-1.

@@

%%10
%\item \label{rfa:enlit:Kvasov2013}
%Kvasov~D.E., Sergeev~Ya.D. Metody lipshicevoj globalnoj optimizacii v zadachax upravleniya~// Avtomatika i telemexanika. 2013. No.~9. S.~3--19.
\item \label{rfa:enlit:Usova2024}
Usova~M.A., Barkalov~K.A. An Algorithm for Finding the Global Extremum of a Partially Defined Function~// Communications in Computer and Information Science. 2024. Vol. 1914. P.~147--161. %DOI: 10.1007/978-3-031-52470-7{\_}13.

@@

\item \label{rfa:enlit:iOptPaper}
Sysoev~A.V., Kozinov~E.A., Barkalov~K.A., Lebedev~I.G., Karchkov~D.A., Rodionov~D.M. Frejmvork metodov intellektualnoj evristicheskoj optimizacii iOpt // V kn.: Su\-per\-kom\-pyu\-ter\-nye dni v Rossii: Trudy mezhdunarodnoj konferencii. 2023. S.~179--185.

\item \label{rfa:enlit:iOptGithub}
Isxodnyj kod frejmvorka iOpt -- URL: https://github.com/aimclub/iOpt (data ob\-ra\-she\-ni\-ya: 26.01.2025).

\item \label{rfa:enlit:iOptDocs}
Dokumentaciya iOpt -- URL: https://iopt.readthedocs.io/ru/latest/ (data ob\-ra\-she\-ni\-ya: 26.01.2025).

@@

\item \label{rfa:enlit:nikitin2022automated}
Filippou~K., Aifantis~G., Papakostas~G.A., Tsekouras~G.E. Structure learning and hy\-per\-pa\-ra\-me\-ter optimization using an automated machine learning ({AutoML}) pipeline // Information. 2023. Vol.~14, No. 4. p.~232.

\item \label{rfa:enlit:FEDOT}
Automated modeling and machine learning framework {FEDOT} -- URL:\\https://github.com/aimclub/FEDOT (data obrashheniya: 25.07.2025).

\item \label{rfa:enlit:MonthlyBeerArticle}
Xu~N. Time Series Analysis on Monthly Beer Production in Australia~// Highlights in Science, Engineering and Technology. 2024. Vol.~94. P.~392--401.% DOI: 10.54097/4z3krj13

@@

\item \label{rfa:renlit:optuna}
Akiba T., Sano S., Yanase T., Ohta T., Koyama M. Optuna: A Next-Generation Hy\-per\-pa\-ra\-me\-ter Optimization Framework // In Proceedings: 25th ACM SIGKDD International Conference on Knowledge Discovery \& Data Mining. 2019. P. 2623--2631.%DOI: 10.1145/3292500.3330701

@@

%\newpage

@@

В качестве второй модельной задачи рассмотрим задачу настройки гиперпараметров методов предсказания значений временного ряда (time series forecasting). Предсказание значений производилось с использованием фреймворка FEDOT [\ref{rfa:rulit:nikitin2022automated}, \ref{rfa:rulit:fedot}]
(фреймворк с открытым исходным кодом для задач автоматизированного моделирования и машинного обучения, AutoML). FEDOT поддерживает полный цикл решения задачи, включающий препроцессинг, выбор модели, настройку, кроссвалидацию и т.д. Фреймворк использует ML-модели, в основном из стандартных библиотек \textit{sklearn}, \textit{statsmodels} и \textit{keras}.

@@

Для сравнения поставленная задача также была решена при помощи известного фреймворка оптимизации гиперпараметров Optuna [\ref{rfa:rulit:optuna}].
Оба решателя были интегрированы в FEDOT и задействованы в вычислительном эксперименте.

@@

\item \label{rfa:rulit:Frazier2018}
Frazier~P.I. A Tutorial on Bayesian Optimization~// arXiv. 2018.\\DOI: 10.48550/arXiv.1807.02811.

@@

%13
\item \label{rfa:rulit:Birect2020}
Paulavi{\v c}ius~R., Sergeyev~Y.D., Kvasov~D.E., {\v Z}ilinskas~J. Globally-biased {BIRECT} al\-go\-rithm with local accelerators for expensive global optimization~// Expert Syst. Appl. 2020. Vol.~144. p.~113052. DOI: 10.1016/j.eswa.2019.113052.

@@

%1
\item \label{rfa:rulit:Audet2022}
Audet~C., Batailly~A., Kojtych~S. Escaping unknown discontinuous regions in blackbox optimization~// SIAM J. Optim. 2022. Vol.~32, No.~3. P.~1843--1870. \\DOI: 10.1137/21m1420915.

%3
\item \label{rfa:rulit:Candelieri2019}
Candelieri~A. Sequential model based optimization of partially defined functions
under unknown constraints~// J. Glob. Optim. 2019. Vol.~79, No.~2. P.~281--303. \\DOI: 10.1007/s10898-019-00860-4.

@@

\item \label{rfa:rulit:Usova2024}
Usova~M.A., Barkalov~K.A. An Algorithm for Finding the Global Extremum of a Partially Defined Function~//
Communications in Computer and Information Science. 2024. Vol. 1914. P.~147--161. DOI: 10.1007/978-3-031-52470-7{\_}13.

@@

\item \label{rfa:rulit:nikitin2022automated}
Filippou~K., Aifantis~G., Papakostas~G.A., Tsekouras~G.E. Structure learning and hy\-per\-pa\-ra\-me\-ter optimization using an automated machine learning ({AutoML}) pipeline // Information. 2023. Vol.~14, No. 4. p.~232.

\item \label{rfa:rulit:fedot}
Automated modeling and machine learning framework {FEDOT} -- URL: \\https://github.com/aimclub/FEDOT (дата обращения: 25.07.2025).

@@

\item \label{rfa:rulit:optuna}
Akiba T., Sano S., Yanase T., Ohta T., Koyama M. Optuna: A Next-Generation Hyperparameter Optimization Framework // In Proceedings: 25th ACM SIGKDD International Conference on Knowledge Discovery \& Data Mining. 2019. P. 2623--2631. DOI: 10.1145/3292500.3330701
